DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-08-04 2015CompleteDerivation/completeDerivation.tex
First pass at kinematic assumption

As we need to describe trajectory, we introduce $\mathcal{M}$ as the manifold that defines space and time. We call $x^i$ the spatial coordinates and $t$ the time coordinate. A trajectory will be given by a set of functions $x^i(s)$ and $t(s)$. What we need to understand is the link between these trajectories and the complete state space of the particles of a classical material.

We first note that the system of units for the trajectories is fully defined by $\mathcal{M}$. Any trajectory can be specified once a set of coordinates $(t, x^i)$ is chosen. Under the kinematic assumption, these units must also be sufficient to define the dynamics. Therefore $\mathcal{M}$ has conceptually two functions: it is the space where the motion unfolds and it is the space that defines the system of units for our states. This already tells us that the our state space is a hyper-surface of $T^*\mathcal{M}$.

@@

Because of the double role of $\mathcal{M}$, we can set half of our state variables to be equal to the space-time variables: $(t, q^i) = (t, x^i)$. This already tells us that the trajectory is differentiable since $q^i(t)$ is differentiable. The four velocity $(u^t, u^i) = (d_s t, d_s x^i)$ is therefore well defined and it can be used to identify trajectories. The magnitude of the four velocity can be rescaled by changing the evolution parameter $s$, and is therefore not physical. Only the direction is physical so the four velocity gives us another $n$ independent variables to identify trajectories. The combined position and velocity, which we call \emph{initial conditions}, identify $2n + 1$ trajectories using a hyper-surface of the tangent bundle $T\mathcal{M}$. This is exactly the dimensionality of the complete state space $\bar{\mathcal{S}} \subset T^*\mathcal{M}$. We cannot add more curves, as a homeomorphism preserve dimensionality, therefore we found the suitable space for our trajectories.

As $\omega(dt,dq^i,dE,dp_i)$ allows us to count possibilities in terms of state variable, we will have a metric $g(dt,dx^i,du^t,du^i)$ that allows us to count the possibilities in terms of initial conditions. Since the both $(dt,dx^i)$ and $(du^t,du^i)$ are four vectors and the metric must be linear, $g$ is a rank two tensor on $\mathcal{M}$. $g$ is also symmetric. Suppose $(t, x^i)$ are orthogonal directions, then a change of unit of one coordinate will only change the component of position and velocity for that coordinate. That is: the component of position and velocity form an independent d.o.f.. For that coordinate system it must be $g(dx^i, du^j)=0=g(dx^j, du^i)$ for $i\neq j$. And the same must hold across temporal and standard d.o.f.. As $g$ is a tensor, the symmetry applies to all choices of coordinates.

@@

\begin{prop}\label{riemannian_manifold}
	The state space $\bar{\mathcal{S}} \subset T^*\mathcal{M}$ is isomorphic to a hyper-surface of the tangent bundle $T\mathcal{M}$ of a pseudo-Riemannian manifold $(\mathcal{M}, g)$.
\end{prop}

@@

\subsection{Scalar and vector potential forces}

>>>>

As we need to describe trajectories, we introduce $\mathcal{M}$ as the manifold that defines space and time. We call $x^i$ the spatial coordinates, $t$ the time coordinate and $x^\alpha = (t, x^i)$ their combination. A trajectory will be given by a set of functions $x^\alpha(s)$. What we need to understand is the link between these trajectories and the complete state space of the particles of a classical material.

We first note that the system of units for the trajectories is fully defined by $\mathcal{M}$. Any trajectory can be specified once a set of coordinates $x^\alpha$ is chosen. Under the kinematic assumption, these units must also be sufficient to define the dynamics. Therefore $\mathcal{M}$ has conceptually two functions: it is the space where the motion unfolds and it is the space that defines the system of units for our states. This already tells us that the our state space is a hyper-surface of $T^*\mathcal{M}$.

@@

Because of the double role of $\mathcal{M}$, we can set half of our state variables to be equal to the space-time variables: $(t, q^i) = x^\alpha$. This already tells us that the trajectory is differentiable since $q^i(t)$ is differentiable. The four velocity $u^\alpha = d_s x^\alpha$ is therefore well defined and it can be used to identify trajectories. The magnitude of the four velocity can be rescaled by changing the evolution parameter, $u^\alpha = d_s x^\alpha = d_s \hat{s} \, d_{\hat{s}} x^\alpha = d_s \hat{s} \, \hat{u}^\alpha$, and is therefore not physical. Only the direction is physical so the four velocity gives us another $n$ independent variables to identify trajectories. The combined position and velocity, which we call \emph{initial conditions}, identify $2n + 1$ trajectories using a hyper-surface of the tangent bundle $T\mathcal{M}$. This is exactly the dimensionality of the complete state space $\bar{\mathcal{S}} \subset T^*\mathcal{M}$. We cannot add more curves, as a homeomorphism preserve dimensionality, therefore we found the suitable space for our trajectories.

@@

\begin{prop}\label{initial_conditions}
	Let $\mathcal{M}$ be the $n+1$ dimensional space-time manifolds over which trajectories are defined. The state space $\bar{\mathcal{S}} \subset T^*\mathcal{M}$ is isomorphic to a hyper-surface of the tangent bundle $T\mathcal{M}$.
\end{prop}

@@

This relationship allows us to write the initial conditions in terms of the state variable. In particular we have $u^\alpha=u^\alpha(t, q^i, E, p_i)$. As the relationship is invertible, these functions must be monotonic in $E$ and $p_i$. Since $\partial_{p_i} \mathcal{H}=d_s q^i= u^i$ and $- \partial_{E} \mathcal{H}= d_s t= u^0$, $\mathcal{H}$ is convex in $E$ and $p_i$. Under such conditions we can define a Legendre transformation $\mathcal{L}=u^\alpha p_\alpha - \mathcal{H}$. Such transformation is the Lagrangian of the system and laws of motions are given by the Euler-Lagrange equation.

\begin{prop}\label{lagrangian_mechanics}
	Let $\mathcal{H}$ be an extended Hamiltonian defined on $\mathbf{T}^*\mathcal{M}$. Under the kinematic assumption, we can define the Legendre transform $\mathcal{L}=u^\alpha p_\alpha - \mathcal{H}$ which we call the \emph{extended Lagrangian}. For the time invariant case, we define the \emph{Lagrangian} $L=v^i p_i - H$.
\end{prop}

Being able to convert the state variable for the particles into initial conditions: we also need to be able to express the distributions of the composite states over position and velocity. To do that, we need to be able to count trajectories in the same way we are able to count states. There must exist a metric $g$, consistent with $\omega$, that allows to calculate the cardinality of possibilities for initial conditions. That is: $\omega(dt,dq^i,dE,dp_i) \propto g(dx^\alpha,du^\alpha)$.

Since the both $dx^\alpha$ and $du^\alpha$ are four vectors and the metric must be linear, $g$ is a rank two tensor on $\mathcal{M}$. This means that a each direction in space-time $dx^\alpha$ is associated with a direction in the space of velocities $du^\alpha$ to form an independent d.o.f. Across independent d.o.f. we have $\omega(dq^i, dp_j) = 0 =\omega(dq^j, dp_i)$. Therefore if $dx^i$ and $dx^j$ do not form a d.o.f., $g(dx^i, du^j) = 0 = g(dx^j, du^i)$: $g$ is reflexive. We also know that $g$ cannot be alternating, that is $g(dx^i, du^i) \neq 0$: a choice of units for $x^i$ is a choice of units for $u^i$ and therefore they cannot belong to different d.o.f. This means $g$ is symmetric. $\mathcal{M}$ is a pseudo-Riemannian manifold.

Since $g$ is symmetric, it can be diagonalized: $\omega= dq^i \wedge dp_i - dt \wedge dE \propto g = dx^i g_{ii} du^i + dt g_{00} du^0$. This tells us $g_{ii} > 0$ and $g_{00} < 0$. We can further change the units along the spatial directions so that $g_{ii}=1$: if we use the same units for the spatial coordinates, each direction will contribute the same number of possibilities. Since time must use a different unit, we set $g_{00}=-c^2$. Therefore, locally $g(dx^i, dx^i) = (dx^i)^2 - c^2 dt^2$ is Minkowskian.

\begin{prop}\label{riemannian_manifold}
	The space-time manifold $\mathcal{M}$ is a pseudo-Riemannian manifold $(\mathcal{M}, g)$ and locally Mikowskian. That is, locally $g = (dx^i)^2 - c^2 dt^2$.
\end{prop}

The metric tensor $g$ and the constant $c$ found here are indeed the ones used in the context of relativity. Yet, they have acquired a different meaning. $g$ does not measure space-time distances: it measure cardinality of possibilities, like $\omega$. Directions in space are orthogonal not because the angle between them is $90^\circ$, but because they identify independent d.o.f.. Space and time, instead, are not independent and the geometry between them is different. In other words: $g$ inherits it's properties from $\omega$.

This also allows to give $c$ a more insightful physical meaning. It is not a velocity: the metric is defined at each point without motion. It is the constant of proportionality that maps a range of initial condition from the time variable $t$ to the space variable $x^i$. In the same way that $1$km would correspond to a thousand times more initial conditions than $1$m, $1$sec corresponds to $299,792,458$ more initial conditions than $1$m. Therefore when we write $x^\alpha = (ct, x^i)$ or $p_\alpha = (- E/c, p_i)$ we are really expressing everything in units that are proportional to the cardinality of state and possibilities. Naturally, these units make the book keeping of number of states much easier which, as we saw, is the only thing that Hamiltonian and Lagrangian mechanics are doing.

This also allows to understand conceptually why the speed of a deterministic and reversible signal cannot be greater than $c$. Consider the trajectory of a point particle. Over an interval $dt$ it will travel $dx$ in space. Since the motion is deterministic and reversible, it will travel through all the states defined in the interval $dx$. It will also travel through every moment in time, which corresponds to $cdt$ possibilities. The number of states that will travel in space cannot exceed the number of states it travels in time as it can only be at one position for each instant. Therefore we have:
\begin{align*}
\frac{|n_x(dx)|}{|n_t(dt)|} &=  \frac{|dx|}{|c dt|} \leq 1 \\
\frac{|dx|}{|dt|} & \leq c \\
\end{align*}
Note, though, that this limit applies to deterministic and reversible motion. It was under that premise that we were able to show that the trajectories are continuous, that we could define $\omega$ and therefore $g$. Non-local quantum effects, such as the ones features in EPR experiments, are not examples of deterministic and reversible motion. Therefore they do not have anything to violate: the maximum speed, simply put, does not apply.

\subsection{Massive particles under scalar and vector potential forces}

The link between $\omega$ and $g$ allows us to constrain the kind of homemorphism between state variables and initial conditions. If we express the possibilities in terms of both state variables and initial conditions we have: $dq^\alpha \wedge dp_\alpha \propto dx^\alpha du_\alpha$. Since  $p_\alpha=p_\alpha(x, u)$, we have $dp_\alpha = m du_\alpha$, where $m$ is the constant of proportionality. We recognize $m$ as the inertial mass. Integrating the expression, we have $p_\alpha = m u_\alpha + A_\alpha(x)$ where $A_\alpha(x)$ is an arbitrary function. This is the most general link between velocity and conjugate momentum, which is the sum of kinetic momentum and a scalar potential.

This constrains the form of the extended Hamiltonian. $\frac{1}{m}g^{\alpha\beta}(p_\beta-A_\beta) = u^\alpha = d_s x^\alpha = d_s q^\alpha = \partial_{p_\alpha} \mathcal{H}$. Integrating we have $\mathcal{H}=\frac{1}{2m}(p_\alpha-A_\alpha(x))g^{\alpha\beta}(p_\beta-A_\beta(x))+V(x)$ where $V(x)$ is an arbitrary function. We can further constrain the Hamiltonian by choosing the natural parameterization for $s$ (also called unit speed or arc length parametrization) for which the modulus of the velocity is constant. We have $u^\alpha u_\alpha = - c^2 = \frac{1}{m^2}(p^\alpha-A^\alpha)(p_\alpha-A_\alpha)$. $\mathcal{H}=\frac{1}{2m}((p_\alpha-A_\alpha(x)g^{\alpha\beta}(p_\beta-A_\beta(x) + m^2 c^2)$. Applying the same procedure in the time independent case yields the Hamiltonian $H=\frac{1}{2m}(p^i-A^i(x))(p_i-A_i(x))+V(x)$.

We recognize the Hamiltonian for massive particles under scalar and vector potential forces.

\begin{prop}\label{kineticHamiltonian}
The extended Hamiltonian for the particles of a classical material under the kinematic assumption is $\mathcal{H}=\frac{1}{2m}((p_\alpha-A_\alpha)g^{\alpha\beta}(p_\beta-A_\beta) + m^2 c^2)$.
\end{prop}

The above discussion gives the inertial mass a new physical meaning: it is the conversion constant that tells how to convert a range of velocity to a range of conjugate momentum with the same number of possibilities. In other words: the more massive the particles, the more states per unit velocity. Intuitively, a more massive body is harder to accelerate because it has to go through more states for the same change in velocity. Note that the Hamiltonian excludes massless particles and this makes absolute sense. A massless particle always has the same velocity, $c$, therefore the trajectory is not enough to tell us the dynamics. The motion of the photon is not enough to tell us its energy or its momentum. Therefore it makes sense that, by assuming kinematic equivalence we implicitly required massive particles.

We have seen that the inertial mass $m$, the invariant speed $c$ and the reduced Plank constant $\hbar$ all serve the same purpose: convert one quantity to another preserving the number of possibilities. $dp=\hbar dk$ converts between ranges of wave number and conjugate momentum. $dx = c dt$ converts between ranges of space and time. $dp = m du$ converts between ranges of velocity and conjugate momentum. They allow to keep proper book keeping when mapping sets of states. There is a difference, though: while $c$ and $\hbar$ are simply determined by the unit system, $m$ also depends on the type of material. That is, different homogeneous materials can have different inertial mass.

Another important thing to notice is that the relationship between conjugate momentum and velocity, $p_\alpha = m u_\alpha + A_\alpha(x)$, includes an arbitrary function. As it is arbitrary, we cannot give $p$ a well defined physical meaning in terms of the space-time trajectory. Suppose, in fact, we changed $A$ and $p$ while leaving $u$ unchanged, the physical description in terms of trajectories remains the same. Therefore we should ask whether any choice of $A$ actually represent a different case leading to different trajectories.

The simplest and more insightful way to study this problem is simply to use $x^\alpha$ and $u_\alpha$ as state variable. They are not going to be conjugates of each other but they do capture the entire physical description. Any change in $A$ that does not affect $x^\alpha$ and $mu_\alpha$ does not change the physics. Expressing the Hamiltonian and the Poisson brackets in terms of the initial conditions we have:
\begin{align*}
\mathcal{H}&=\frac{1}{2m}(mu_\alpha mu^\alpha + m^2 c^2) \\
\{x^\alpha, x^\beta\} &= 0 \\
\{x^\alpha, mu_\beta\} &= \delta^\alpha_\beta \\
\{mu_\alpha, mu_\beta\} &= \partial_\alpha A_\beta - \partial_\beta A_\alpha = F_{\alpha\beta}
\end{align*}
Note how the Poisson bracket between kinetic momentum is the force tensor. The weaker the force, the more $(x^i, mu^i)$ become independent d.o.f. The force tensor, then, tells how much $(x^i, mu^i)$ fail to be independent d.o.f.. With these expressions, the commutator between any function of position and kinetic momentum can now be calculated. For example:
\begin{align*}
d_s x^\alpha &= \{x^\alpha, \mathcal{H}\} = u^\alpha \\
d_s (mu^\alpha) &= \{mu^\alpha, \mathcal{H}\} = F^{\alpha\beta} u_\beta
\end{align*}
where we recognize the covariant form of the Lorentz force.

Any change of $A$ that leaves $F_{\alpha\beta}$ unchanged does not affect the motion of any function of position and velocity. Such transformation is what we call a gauge transformation.

As we mentioned before, as the extended Hamiltonian is quadratic in the energy, negative energy states are possible. As we fixed the natural parametrization, $u^\alpha u_\alpha = - c^2$, there are two cases: the parametrization is aligned with time, therefore $s=\tau$ where $\tau$ is proper time, or the parametrization is anti-aligned with time, therefore $s=-\tau$. The trajectory velocity $u^\alpha= \pm d_\tau x^\alpha$ is plus or minus the physical velocity depending on whether it's a standard or anti-particle state. Expanding the equation for the force, we have:
\begin{align*}
\pm d_\tau (\pm m d_\tau x^\alpha) &= m d_\tau^2 x^\alpha = \pm  F^\alpha_\beta d_\tau x^\beta
\end{align*}
Therefore the force act on anti-particle states in the opposite way: they act as they have opposite charge. Note that nothing physical is actually going backwards in time, just the parametrization.

In this setting, charge conjugation $C$ is simply changing the parametrization from $s$ to $-s$, following the trajectories in the opposite way. This will permute each particle anti-state with the corresponding standard state and vice-versa. But if we apply parity $P$ and time reversal $T$, changing $x^\alpha$ to $-x^\alpha$, we would have the same net effect: the parametrization stays the same, but it now moves backwards in space and time. As they have the same net effect, if we apply all three we go back where we started. This gives us a very intuitive classical particle mechanics analogue for the $CPT$ theorem.

To sum up, the kinematic assumption has more inherent consequences one would expect at first. It means the particles are identified by position and velocity, the system of equation is second order. It links the space of initial conditions on $T\mathcal{M}$ with the space of states on $T^*\mathcal{M}$. Such link allows state space metric $\omega$ to induce the space-time metric $g$. It allows us to define Lagrangians. It constrains the motion to massive particles under scalar and vector potential forces. Moreover, it allowed to re-understand basic concepts, like invariant speed $c$, the inertial mass $m$ and the force tensor $F_{\alpha\beta}$ in terms of state counting and d.o.f. independence.

We now have all the building block of classical mechanics, the consequences of three seemingly simple assumption. But that's not all: we have placed ourselves in a very good position to move forward to quantum systems as well by using the new physical insights we have gained.
